---
title: "Count 18 Mash - IndRanBayes - EE"
author: "Yuxin Zou"
date: 2018-4-16
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

```{r, echo=FALSE}
# TEMPORARY.
knitr::opts_chunk$set(eval = TRUE)
```

```{r packages}
library(limma); library(mashr); library(mclust); library(plyr);
library(flashr); library(colorRamps); library(corrplot)
```

Implement `mash` on John's data at time point 18.

```{r data}
data = readRDS('../output/Count_IndRan_limmaBayes.rds')
# Time 18
Eighteen = grepl('t18', colnames(data$beta))

beta_class = data$beta[, Eighteen]
sebeta_class = data$sebeta[, Eighteen]
colnames(beta_class) = colnames(sebeta_class) = gsub( "\\..*$", "", colnames(beta_class))
```

# Mash

Set mash data object: the standard errors in the data are corresponding to t statistics.

```{r mash.data}
mash.data = mash_set_data(Bhat = beta_class, Shat = sebeta_class, df = data$df[1])
L = diag(ncol(beta_class))
# the first col is none
L[,1] = -1
L = L[2:9,]
mash.data.diff = mash_set_data_contrast(mash.data, L)
```

Top genes:
```{r top genes}
# find strong genes
m.1by1 = mash_1by1(mash.data.diff, alpha=0)
strong = get_significant_results(m.1by1)
Z = mash.data.diff$Bhat/mash.data.diff$Shat
Z.strong = Z[strong,]
# center
Z.center = apply(Z.strong, 2, function(x) x - mean(x))
```

## Estimate covariance structures:

Data Driven:

Flash:
```{r flash}
flash.data = flash_set_data(Z.center)
fmodel = flash(flash.data, greedy = TRUE, backfit = TRUE)
flash_get_pve(fmodel)
```
```{r flash plot}
factors = flash_get_ldf(fmodel)$f
row.names(factors) = colnames(beta_class)[2:9]
pve.order = order(flash_get_pve(fmodel), decreasing = TRUE)

par(mfrow=c(2,3))
for(i in pve.order){
  barplot(factors[,i], main=paste0('Factor ',i, ' pve= ', round(flash_get_pve(fmodel)[i],3)), las=2, cex.names = 0.7)
}
par(mfrow=c(1,1))
```

`flash` on the loading:
```{r flashLoading}
loading = fmodel$EL[,1:5]
colnames(loading) = paste0('Factor',seq(1,5))
flash.loading = flash_set_data(loading)
flmodel = flash(flash.loading, greedy = TRUE, backfit = TRUE)
```

The flash prefers the rank 0 model. There is no hidden structure in the loading matrix.

Cluster loadings:
```{r cluster loading}
mod = Mclust(loading)
summary(mod$BIC)
```

Using clustering result to fit `mash`:

$$l_{i}\sim \sum_{j=1}^{m}N(\mu_{j}, \Sigma_{j})$$
We estimate the covariance as $F(\Sigma_j + \mu_{j}\mu_{j}')F'$.
```{r cov from flash}
U_list = alply(mod$parameters$variance$sigma,3)
mu_list = alply(mod$parameters$mean,2)
ll = list()
for (i in 1:length(U_list)){
  ll[[i]] = U_list[[i]] + mu_list[[i]] %*% t(mu_list[[i]])
}

Factors = fmodel$EF[,1:5]
U.loading = lapply(ll, function(U){Factors %*% (U %*% t(Factors))})
names(U.loading) = paste0('Load', "_", (1:length(U.loading)))

# rank 1
Flash_res = flash_get_lf(fmodel)
U.Flash = c(mashr::cov_from_factors(t(as.matrix(factors)), "Flash"), 
            list("tFlash" = t(Flash_res) %*% Flash_res / nrow(Z.center)))
```

PCA:
```{r pca}
U.pca = cov_pca(mash_set_data(Z.center), 4)
```

Canonical
```{r Canonical}
U.c = cov_canonical(mash_set_data(Z.center))
```

Extreme Deconvolution
```{r ED, eval=FALSE}
U.dd = c(U.pca, U.loading, U.Flash, list('XX' = t(Z.center) %*% Z.center / nrow(Z.center)))

mash.data.ed = mash.data.diff
mash.data.ed$Bhat = mash.data.diff$Bhat[strong,]
mash.data.ed$Shat = mash.data.diff$Shat[strong,]
mash.data.ed$Shat_alpha = mash.data.diff$Shat_alpha[strong,]
saveRDS(cov_ed(mash.data.ed, U.dd), '../output/Count18_IndRanBayes_MashEE_Cov.rds')
```

## Mash model:
```{r, echo=FALSE}
U.ed = readRDS('../output/Count18_IndRanBayes_MashEE_Cov.rds')
```

```{r mash, eval=FALSE}
saveRDS(mash(mash.data.diff, c(U.c, U.ed), algorithm.version = 'R'), '../output/Count18_IndRanBayes_mash_model_EE.rds') 
```

# Result
```{r}
mash.model = readRDS('../output/Count18_IndRanBayes_mash_model_EE.rds')
```

The log-likelihood of fit is
```{r loglik} 
get_loglik(mash.model)
```
Here is a plot of weights learned:
```{r estimated weight}
options(repr.plot.width=12, repr.plot.height=4)
barplot(get_estimated_pi(mash.model), las = 2, cex.names = 0.7)
```

Check Load_6 covariance matrix

```{r Load6}
x           <- mash.model$fitted_g$Ulist[["ED_Load_6"]]
colnames(x) <- colnames(beta_class)[2:9]
rownames(x) <- colnames(x)
corrplot(x, method='color', cl.lim=c(0,1), type='upper', addCoef.col = "black", tl.col="black", tl.srt=45)
```

```{r Load6 pca}
layout(matrix(c(1,2,3,4), 2, 2, byrow=TRUE))
svd.out = svd(mash.model$fitted_g$Ulist[["ED_Load_6"]])
v = svd.out$v
colnames(v) = colnames(beta_class)[-1]
rownames(v) = colnames(v)
options(repr.plot.width=10, repr.plot.height=5)
for (j in 1:4)
  barplot(v[,j]/v[,j][which.max(abs(v[,j]))], cex.names = 0.7,
          las = 2, main = paste0("EigenVector ", j, " for Load 6"))
```

Check Load_8 covariance matrix

```{r Load8}
x           <- mash.model$fitted_g$Ulist[["ED_Load_8"]]
colnames(x) <- colnames(beta_class)[2:9]
rownames(x) <- colnames(x)
corrplot(x, method='color', cl.lim=c(0,1), type='upper', addCoef.col = "black", tl.col="black", tl.srt=45)
```

```{r Load8 pca}
layout(matrix(c(1,2,3,4), 2, 2, byrow=TRUE))
svd.out = svd(mash.model$fitted_g$Ulist[["ED_Load_8"]])
v = svd.out$v
colnames(v) = colnames(beta_class)[-1]
rownames(v) = colnames(v)
options(repr.plot.width=10, repr.plot.height=5)
for (j in 1:4)
  barplot(v[,j]/v[,j][which.max(abs(v[,j]))], cex.names = 0.7,
          las = 2, main = paste0("EigenVector ", j, " for Load 8"))
```

Both matrices represent that deviations are shared in sign among conditions. The deviations in Yers and Salm are strongest, but weaker in BCG and RV+.

There are `r length(get_significant_results(mash.model))` diferentially expressed genes.

Check pairwise sharing by magnitude and sign:
```{r}
x = get_pairwise_sharing(mash.model)
x[x > 1]    <- 1
x[x < -1]   <- -1
colnames(x) <- colnames(beta_class)[-1]
rownames(x) <- colnames(x)
corrplot.mixed(x, tl.pos="d",upper='color', cl.lim=c(0,1), upper.col=colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF",
                               "#E0F3F8","#91BFDB","#4575B4")))(40),
               tl.cex=1.2)
```

Check pairwise sharing by sign:
```{r}
x = get_pairwise_sharing(mash.model, factor=0)
x[x > 1]    <- 1
x[x < -1]   <- -1
colnames(x) <- colnames(beta_class)[-1]
rownames(x) <- colnames(x)
corrplot.mixed(x, tl.pos="d",upper='color', cl.lim=c(0,1), upper.col=colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF",
                               "#E0F3F8","#91BFDB","#4575B4")))(40),
               tl.cex=1.2)
```

# Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
