---
title: "Count - Individual Fix Effect - Bayes"
author: "Yuxin Zou"
date: 2018-4-16
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the R version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r R-version, echo=FALSE, results='asis'}
```

```{r, echo=FALSE}
# TEMPORARY.
knitr::opts_chunk$set(eval = TRUE)
```

The type of bacteria (bact), and the time post-infection (time) are treated as fixed effects. The individual (ind) is treated as random effect.

The covariates RIN (rin) and extraction batch (extr) are included in the model to correct for these confounding effects.

```{r packages}
library("limma")
library("edgeR")
library('lme4')
library('car')
```

Read in the data of counts (row = sample, column = feature).

```{r data}
full <- read.table("../data/counts_per_sample.txt", header = TRUE,
                   stringsAsFactors = FALSE)
full <- full[order(full$dir), ]
rownames(full) <- paste(full$dir, full$ind, full$bact, full$time, sep = ".")
counts <- t(full[, grep("ENSG", colnames(full))])
# Filter lowly expressed genes
counts <- counts[rowSums(cpm(counts) > 1) > 6, ]
```

The first `r sum(!grepl("ENSG", colnames(full)))` columns are meta-data about each sample, and the remaining are the gene counts.

Only the `r nrow(counts)` genes which have at least one read per million (pre-normalization with TMM) in at least six of the `r nrow(full)` samples are kept.

Prepare the factors for use in the linear model.

```{r annotation}
groups <- full[, c("ind", "bact", "time", "extr", "rin")]
groups$bact <- gsub("\\+", "plus", groups$bact)
groups$ind <- factor(groups$ind)
groups$bact <- factor(groups$bact, levels = c("none", "Rv", "Rvplus", "GC",
                                              "BCG", "Smeg", "Yers", "Salm",
                                              "Staph"))
groups$time <- factor(groups$time, levels = c(4, 18, 48))
groups$extr <- factor(groups$extr)
head(groups)
```

* Create the model using a treatment parametrization
```{r model}
contrasts(groups$extr) = contr.Sum(levels(groups$extr))
contrasts(groups$ind) = contr.Sum(levels(groups$ind))
design <- model.matrix(~ 0 + bact:time, data = groups)
batch <- model.matrix(~ ind + extr + rin, data = groups)
design <- cbind(design, batch[,-1])
# Clean up names of design matrix
colnames(design) <- gsub("ind", "", colnames(design))
colnames(design) <- gsub("time", "t", colnames(design))
colnames(design) <- gsub("bact", "", colnames(design))
colnames(design) <- gsub(":", ".", colnames(design))
# Remove Staph 48 hr timepoint that does not exist
design <- design[, -which(colnames(design) == "Staph.t48")]
print(colnames(design))
```

* Normalize with TMM (`calcNormFactors`)

```{r TMM}
y <- DGEList(counts)
y <- calcNormFactors(y)
```

* Transfer with `voom`
```{r voom}
v = voom(y, design)
```

* limma
```{r limma}
fit = lmFit(v, design)
fitB = eBayes(fit)
```

```{r assign}
beta_class = fitB$coefficients[,1:26]
sebeta_class = fitB$stdev.unscaled[, 1:26] * sqrt(fitB$s2.post)
t_class = fitB$t[, 1:26]
df_class = fitB$df.total
```

```{r save}
saveRDS(list(beta = beta_class, sebeta = sebeta_class, t = t_class, df = df_class, groups = groups), '../output/Count_IndFix_limmaBayes.rds')
```

# Session information

<!-- Insert the session information into the document -->
```{r session-info}
```

