---
title: "Count 18 Mash - IndRanBayes - EE - Median"
author: "Yuxin Zou"
date: 2018-11-7
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

```{r packages}
library(limma); library(mashr); library(mclust); library(plyr);
library(flashr); library(colorRamps); library(corrplot)
```

Implement `mash` on John's data at time point 18.

```{r data}
data = readRDS('../output/Count_IndRan_limmaBayes.rds')
# Time 18
Eighteen = grepl('t18', colnames(data$beta))

beta_class = data$beta[, Eighteen]
sebeta_class = data$sebeta[, Eighteen]
colnames(beta_class) = colnames(sebeta_class) = gsub( "\\..*$", "", colnames(beta_class))
```

# Mash

Set mash data object: the standard errors in the data are corresponding to t statistics.

```{r mash.data}
mash.data = mash_set_data(Bhat = beta_class, Shat = sebeta_class, df = data$df[1])
delta = t(apply(mash.data$Bhat, 1, function(x) x - median(x)))
mash.data.diff = mash_set_data(delta, Shat = mash.data$Shat)
```

Top genes:
```{r top genes}
# find strong genes
m.1by1 = mash_1by1(mash.data.diff, alpha=0)
strong = get_significant_results(m.1by1)
Z = mash.data.diff$Bhat/mash.data.diff$Shat
Z.strong = Z[strong,]
# center
Z.center = apply(Z.strong, 2, function(x) x - mean(x))
```

## Estimate covariance structures:

Data Driven:

Flash:
```{r flash}
fmodel = flash(Z.center, var_type = 'constant', greedy = TRUE, backfit = TRUE, verbose = FALSE)
```

```{r flash plot}
factors = fmodel$ldf$f
row.names(factors) = colnames(mash.data.diff$Bhat)
pve.order = order(fmodel$pve, decreasing = TRUE)
#png('~/Documents/GitHub/MASH_baseline/figures/John18_flash_Median_Factors.png', width = 700, height = 500)
par(mfrow=c(1,3))
for(i in pve.order){
  barplot(factors[,i], main=paste0('Factor ',i, ' PVE = ', round(fmodel$pve[i],3)), las=2, cex.names = 0.7)
}
par(mfrow=c(1,1))
#dev.off()
```

`flash` on the loading:
```{r flashLoading}
loading = fmodel$ldf$l
colnames(loading) = paste0('Factor',seq(1,3))
flmodel = flash(loading, greedy = TRUE, backfit = TRUE, var_type = 'constant', verbose = FALSE)
flactors = flmodel$ldf$f
pve.order = order(flmodel$pve, decreasing = TRUE)
# png('~/Documents/GitHub/MASH_baseline/figures/John18_flash_Median_Factors.png', width = 700, height = 500)
par(mfrow=c(1,3))
for(i in pve.order){
  barplot(flactors[,i], main=paste0('Factor ',i, ' PVE = ', round(flmodel$pve[i],3)), las=2, cex.names = 0.7)
}
par(mfrow=c(1,1))
#dev.off()
```

```{r}
flFactors = fmodel$ldf$f %*% flmodel$ldf$f
par(mfrow=c(1,3))
for(i in 1:3){
  barplot(flFactors[,i], main=paste0('Factor ',i), las=2, cex.names = 0.7)
}
par(mfrow=c(1,1))
```

Cluster loadings:
```{r cluster loading}
mod = Mclust(loading, 1:20)
summary(mod$BIC)
```

```{r cov from flash}
# rank 1
Flash_res = fmodel$fitted_values
U.Flash = c(mashr::cov_from_factors(t(cbind(flFactors, factors)), "Flash"), 
            list("tFlash" = t(Flash_res) %*% Flash_res / nrow(Z.center)))
```

PCA:
```{r pca}
U.pca = cov_pca(mash_set_data(Z.center), 4)
```

Canonical
```{r Canonical}
U.c = cov_canonical(mash_set_data(Z.center))
```

Extreme Deconvolution
```{r ED, eval=FALSE}
U.dd = c(U.pca, U.Flash, list('XX' = t(Z.center) %*% Z.center / nrow(Z.center)))

saveRDS(cov_ed(mash.data.diff, U.dd, subset = strong), '../output/Count18_IndRanBayes_MashEE_Cov_NOMclust_Median.rds')
```

## Mash model:
```{r}
U.ed = readRDS('../output/Count18_IndRanBayes_MashEE_Cov_NOMclust_Median.rds')
```

```{r mash, eval=FALSE}
saveRDS(mash(mash.data.diff, c(U.c, U.ed), algorithm.version = 'R'), '../output/Count18_IndRanBayes_mash_model_EE_NOMclust_Median.rds') 
```

# Result
```{r}
mash.model = readRDS('../output/Count18_IndRanBayes_mash_model_EE_NOMclust_Median.rds')
```

The log-likelihood of fit is
```{r loglik} 
get_loglik(mash.model)
```
Here is a plot of weights learned:
```{r estimated weight}
options(repr.plot.width=12, repr.plot.height=4)
barplot(get_estimated_pi(mash.model), las = 2, cex.names = 0.7)
```

Check Flash 4 covariance matrix

```{r Load6}
x           <- mash.model$fitted_g$Ulist[["ED_Flash_4"]]
colnames(x) <- colnames(mash.data.diff$Bhat)
rownames(x) <- colnames(x)
png('~/Documents/GitHub/MASH_baseline/figures/John18CovMedianFactor4.png', width = 500, height=500)
corrplot(x, method='color', cl.lim=c(-1,1), type='upper', addCoef.col = "black", tl.col="black", tl.srt=45)
dev.off()
```

Both matrices represent that deviations are shared in sign among conditions. The deviations in Yers and Salm are strongest, but weaker in BCG and RV+.

There are `r length(get_significant_results(mash.model))` diferentially expressed genes.

Check pairwise sharing by magnitude and sign:
```{r}
x = get_pairwise_sharing(mash.model)
x[x > 1]    <- 1
x[x < -1]   <- -1
colnames(x) <- colnames(beta_class)
rownames(x) <- colnames(x)
# png(filename = '~/Documents/GitHub/MASH_baseline/figures/John18PS_Mag.png', width=500, height = 500)
corrplot.mixed(x, tl.pos="d",upper='color', cl.lim=c(0,1), upper.col=colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF",
                               "#E0F3F8","#91BFDB","#4575B4")))(40),
               tl.cex=1.2, title = 'Pairwise Sharing by Magnitude', mar = c(4,0,4,0))
# dev.off()
```

Check pairwise sharing by sign:
```{r}
x = get_pairwise_sharing(mash.model, factor=0)
x[x > 1]    <- 1
x[x < -1]   <- -1
colnames(x) <- colnames(beta_class)[-1]
rownames(x) <- colnames(x)
png(filename = '~/Documents/GitHub/MASH_baseline/figures/John18PS_Median_Sign.png', width=500, height = 500)
corrplot.mixed(x, tl.pos="d",upper='color', cl.lim=c(0,1), upper.col=colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF",
                               "#E0F3F8","#91BFDB","#4575B4")))(40),
               tl.cex=1.2, title = 'Pairwise Sharing by Sign', mar = c(4,0,4,0))
dev.off()
```

Sample from posterior:

```{r}
mash.model.sample = mash.model
mash.model.sample$result = mash_compute_posterior_matrices(mash.model, mash.data.diff, algorithm.version = 'R', posterior_samples = 500)
saveRDS(mash.model.sample, '../output/Count18_IndRanBayes_mash_model_EE_NOMclust_Median_samples.rds')
```

```{r}
mash.model.sample = readRDS('../output/Count18_IndRanBayes_mash_model_EE_NOMclust_Median_samples.rds')
```

Propotion of gene sharing sign
```{r}
x = get_pairwise_sharing_from_samples(mash.model.sample, factor = 0)
colnames(x) <- colnames(beta_class)
rownames(x) <- colnames(x)
png(filename = '~/Documents/GitHub/MASH_baseline/figures/John18PS_Median_Sample_Sig_Sign.png', width=500, height = 500)
corrplot(x, method='color', cl.lim=c(0,1), type='upper', addCoef.col = "black", tl.col="black", tl.srt=45, title = 'Pairwise Sharing by Sign', mar = c(4,0,4,0), col=colorRampPalette(c("blue","white","red"))(200))
dev.off()
```

Propotion of gene sharing magnitude
```{r}
x = get_pairwise_sharing_from_samples(mash.model.sample, factor = 0.5)
colnames(x) <- colnames(beta_class)
rownames(x) <- colnames(x)
png(filename = '~/Documents/GitHub/MASH_baseline/figures/John18PS_Median_Sample_Sig_Mag.png', width=500, height = 500)
corrplot(x, method='color', cl.lim=c(0,1), type='upper', addCoef.col = "black", tl.col="black", tl.srt=45, title = 'Pairwise Sharing by Magnitude', mar = c(4,0,4,0), col=colorRampPalette(c("blue","white","red"))(200))
dev.off()
```
# Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
