---
title: "Expression 18 Mash - EE - joint mean MASH model"
author: "Yuxin Zou"
date: "3/13/2021"
output: 
  html_document:
    code_folding: hide
---

We check the result of mash model using jointly estimated group mean.

```{r}
library(mashr);library(colorRamps); library(corrplot)
library(ggplot2); library(ggpubr)
data = readRDS('../output/Expression_john_joint.rds')
beta_class = data$beta
sebeta_class = data$sebeta
mash.data = mash_set_data(Bhat = beta_class, Shat = sebeta_class)
mash.data.L = mash_update_data(mash.data, ref = 'none')
m.model = readRDS('../output/Expression18_jointmean_mash_model.rds')
```

There are `r length(get_significant_results(m.model))` DE genes.

## Patterns

Here is a plot of weights learned:
```{r}
options(repr.plot.width=12, repr.plot.height=4)
barplot(get_estimated_pi(m.model), las = 2, cex.names = 0.7)
```
Check tFlash covariance matrix

```{r tFlash}
x           <- m.model$fitted_g$Ulist[["ED_tFlash"]]
name = colnames(m.model$result$PosteriorMean)
name = gsub('-none', '', name)
colnames(x) <- name
rownames(x) <- colnames(x)
corrplot(x, method='color', cl.lim=c(0,1), type='upper', addCoef.col = "black", tl.col="black", tl.srt=45, tl.cex = 0.7)
```

```{r tFLASH pca}
layout(matrix(c(1,2,3,4), 2, 2, byrow=TRUE))
svd.out = svd(m.model$fitted_g$Ulist[["ED_tFlash"]])
v = svd.out$v
name = colnames(m.model$result$PosteriorMean)
name = gsub('-none', '', name)
colnames(v) = name
rownames(v) = colnames(v)
options(repr.plot.width=10, repr.plot.height=5)
for (j in 1:4)
  barplot(v[,j]/v[,j][which.max(abs(v[,j]))], cex.names = 0.7,
          las = 2, main = paste0("EigenVector ", j, " for ED_tFlash"))
```

```{r, eval=FALSE}
for(i in 1:3){
  png(paste0('../output/Expression18_jointmean_mash_model_tFlash_v',i,'.png'))
  svd.out = svd(m.model$fitted_g$Ulist[["ED_tFlash"]])
  v = svd.out$v
  name = colnames(m.model$result$PosteriorMean)
  name = gsub('-none', '', name)
  colnames(v) = name
  rownames(v) = colnames(v)
  barplot(v[,i]/v[,i][which.max(abs(v[,i]))], cex.names = 0.7, las = 2, 
        main = paste0("EigenVector ", i, " pve=", round(svd.out$d[i]/sum(svd.out$d), 2)))
  dev.off()
}
```

Check XtX covariance matrix

```{r XX}
x           <- m.model$fitted_g$Ulist[["ED_XX"]]
name = colnames(m.model$result$PosteriorMean)
name = gsub('-none', '', name)
colnames(x) <- name
rownames(x) <- colnames(x)
corrplot(x, method='color', cl.lim=c(0,1), type='upper', addCoef.col = "black", tl.col="black", tl.srt=45, tl.cex = 0.7)
```

```{r XX pca}
layout(matrix(c(1,2,3,4), 2, 2, byrow=TRUE))
svd.out = svd(m.model$fitted_g$Ulist[["ED_XX"]])
v = svd.out$v
name = colnames(m.model$result$PosteriorMean)
name = gsub('-none', '', name)
colnames(v) = name
rownames(v) = colnames(v)
options(repr.plot.width=10, repr.plot.height=5)
for (j in 1:4)
  barplot(v[,j]/v[,j][which.max(abs(v[,j]))], cex.names = 0.7,
          las = 2, main = paste0("EigenVector ", j, " for ED_XX"))
```

Check PCA1 covariance matrix

```{r PCA1}
x           <- m.model$fitted_g$Ulist[["ED_PCA_1"]]
name = colnames(m.model$result$PosteriorMean)
name = gsub('-none', '', name)
colnames(x) <- name
rownames(x) <- colnames(x)
corrplot(x, method='color', cl.lim=c(0,1), type='upper', addCoef.col = "black", tl.col="black", tl.srt=45, tl.cex = 0.7)
```

```{r PCA1 pca}
svd.out = svd(m.model$fitted_g$Ulist[["ED_PCA_1"]])
v = svd.out$v
name = colnames(m.model$result$PosteriorMean)
name = gsub('-none', '', name)
colnames(v) = name
rownames(v) = colnames(v)
options(repr.plot.width=10, repr.plot.height=5)
for (j in 1:1)
  barplot(v[,j]/v[,j][which.max(abs(v[,j]))], cex.names = 0.7,
          las = 2, main = paste0("EigenVector ", j, " for ED_PCA_1"))
```

Check Flash3 covariance matrix

```{r Flash3}
x           <- m.model$fitted_g$Ulist[["ED_Flash_3"]]
name = colnames(m.model$result$PosteriorMean)
name = gsub('-none', '', name)
colnames(x) <- name
rownames(x) <- colnames(x)
corrplot(x, method='color', cl.lim=c(-1,1), type='upper', addCoef.col = "black", tl.col="black", tl.srt=45, tl.cex = 0.7)
```


Check pairwise sharing by magnitude and sign:
```{r sharingMagSign}
x = get_pairwise_sharing(m.model)
x[x > 1]    <- 1
x[x < -1]   <- -1
name = colnames(m.model$result$PosteriorMean)
name = gsub('-none', '', name)
colnames(x) <- name
rownames(x) <- colnames(x)
corrplot.mixed(x, tl.pos="d",upper='color', cl.lim=c(0,1), upper.col=colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF",
                               "#E0F3F8","#91BFDB","#4575B4")))(40),
               tl.cex=1.2)
```

Check pairwise sharing by magnitude:
```{r shareMag}
x = get_pairwise_sharing(m.model, FUN = abs)
x[x > 1]    <- 1
x[x < -1]   <- -1
name = colnames(m.model$result$PosteriorMean)
name = gsub('-none', '', name)
colnames(x) <- name
rownames(x) <- colnames(x)
corrplot.mixed(x, tl.pos="d",upper='color', cl.lim=c(0,1), upper.col=colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF",
                               "#E0F3F8","#91BFDB","#4575B4")))(40),
               tl.cex=1.2)
```

Check pairwise sharing by sign:
```{r shareSign}
x = get_pairwise_sharing(m.model, factor=0)
x[x > 1]    <- 1
x[x < -1]   <- -1
name = colnames(m.model$result$PosteriorMean)
name = gsub('-none', '', name)
colnames(x) <- name
rownames(x) <- colnames(x)
corrplot.mixed(x, tl.pos="d",upper='color', cl.lim=c(0,1), upper.col=colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF",
                               "#E0F3F8","#91BFDB","#4575B4")))(40),
               tl.cex=1.2)
```

## Genes

Yers-Salm cluster in Cormotif. All effects in mash.

```{r YSgene1, fig.height=5, fig.width=10}
# before 11952
gene1177 = data.frame(mash.data.L$Bhat[1177,])
colnames(gene1177) = 'Deviations'
gene1177$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene1177$se = mash.data.L$Shat[1177,]
gene1177$Deviations[gene1177$se == 0] = NA
# after
gene1177.post = data.frame(m.model$result$PosteriorMean[1177,])
colnames(gene1177.post) = 'Deviations'
gene1177.post$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene1177.post$se = m.model$result$PosteriorSD[1177,]

p.orig = ggplot(gene1177, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + ylim(-0.5,3) +
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('CDC14A original log 2 fold change' )

p.post = ggplot(gene1177.post, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + ylim(-0.5,3) +
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('CDC14A mash')

ggarrange(p.orig, p.post, ncol=2, nrow=1, heights = c(5, 5), widths = c(5,5))
```

Yers-Salm cluster in Cormotif. Yers-Salm in mash. 

```{r YSgene2, fig.height=5, fig.width=10}
# before
gene7637 = data.frame(mash.data.L$Bhat[7637,])
colnames(gene7637) = 'Deviations'
gene7637$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene7637$se = mash.data.L$Shat[7637,]
gene7637$Deviations[gene7637$se == 0] = NA
# after
gene7637.post = data.frame(m.model$result$PosteriorMean[7637,])
colnames(gene7637.post) = 'Deviations'
gene7637.post$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene7637.post$se = m.model$result$PosteriorSD[7637,]

p.orig = ggplot(gene7637, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + ylim(-8,3) +
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('CD1E original log 2 fold change' )

p.post = ggplot(gene7637.post, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + ylim(-8, 3) +
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('CD1E mash')

ggarrange(p.orig, p.post, ncol=2, nrow=1, heights = c(5, 5), widths = c(5,5))
```

All in Cormofit. All in mash.
```{r allgene1, fig.height=5, fig.width=10}
# before
gene11445 = data.frame(mash.data.L$Bhat[11445,])
colnames(gene11445) = 'Deviations'
gene11445$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene11445$se = mash.data.L$Shat[11445,]
gene11445$Deviations[gene11445$se == 0] = NA
# after
gene11445.post = data.frame(m.model$result$PosteriorMean[11445,])
colnames(gene11445.post) = 'Deviations'
gene11445.post$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene11445.post$se = m.model$result$PosteriorSD[11445,]

p.orig = ggplot(gene11445, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + ylim(0,4) +
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) +
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('HRH1 original log 2 fold change' )

p.post = ggplot(gene11445.post, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + ylim(0, 4) +
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('HRH1 mash')

ggarrange(p.orig, p.post, ncol=2, nrow=1, heights = c(5, 5), widths = c(5,5))
```

All in Cormofit. All in mash.
```{r allgene2, fig.width=10, fig.height=5}
# before
gene2481 = data.frame(mash.data.L$Bhat[2481,])
colnames(gene2481) = 'Deviations'
gene2481$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene2481$se = mash.data.L$Shat[2481,]
gene2481$Deviations[gene2481$se == 0] = NA
# after
gene2481.post = data.frame(m.model$result$PosteriorMean[2481,])
colnames(gene2481.post) = 'Deviations'
gene2481.post$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene2481.post$se = m.model$result$PosteriorSD[2481,]

p.orig = ggplot(gene2481, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + 
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) +
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('ZNF175 original log 2 fold change' )

p.post = ggplot(gene2481.post, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + 
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('ZNF175 mash')

ggarrange(p.orig, p.post, ncol=2, nrow=1, heights = c(5, 5), widths = c(5,5))
```

All in Cormofit. None in mash.

```{r allgene3, fig.height=5, fig.width=10}
# before 1535
gene1535 = data.frame(mash.data.L$Bhat[1535,])
colnames(gene1535) = 'Deviations'
gene1535$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene1535$se = mash.data.L$Shat[1535,]
gene1535$Deviations[gene1535$se == 0] = NA
# after
gene1535.post = data.frame(m.model$result$PosteriorMean[1535,])
colnames(gene1535.post) = 'Deviations'
gene1535.post$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene1535.post$se = m.model$result$PosteriorSD[1535,]

p.orig = ggplot(gene1535, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + ylim(-8,3) +
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('FCGBP original log 2 fold change' )

p.post = ggplot(gene1535.post, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + ylim(-8, 3) +
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('FCGBP mash')

ggarrange(p.orig, p.post, ncol=2, nrow=1, heights = c(5, 5), widths = c(5,5))
```

Non-DE in Cormotif. Strong in Smeg in mash.
```{r Smeggene1, fig.width=10, fig.height=5}
# before
gene5252 = data.frame(mash.data.L$Bhat[5252,])
colnames(gene5252) = 'Deviations'
gene5252$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene5252$se = mash.data.L$Shat[5252,]
gene5252$Deviations[gene5252$se == 0] = NA
# after
gene5252.post = data.frame(m.model$result$PosteriorMean[5252,])
colnames(gene5252.post) = 'Deviations'
gene5252.post$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene5252.post$se = m.model$result$PosteriorSD[5252,]

p.orig = ggplot(gene5252, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + 
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) +
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('RSAD2 original log 2 fold change' )

p.post = ggplot(gene5252.post, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + 
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('RSAD2 mash')

ggarrange(p.orig, p.post, ncol=2, nrow=1, heights = c(5, 5), widths = c(5,5))
```

Non-DE in Cormotif. Strong in Smeg in mash.
```{r Smeggene2, fig.width=10, fig.height=5}
# before
gene12428 = data.frame(mash.data.L$Bhat[12428,])
colnames(gene12428) = 'Deviations'
gene12428$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene12428$se = mash.data.L$Shat[12428,]
gene12428$Deviations[gene12428$se == 0] = NA
# after
gene12428.post = data.frame(m.model$result$PosteriorMean[12428,])
colnames(gene12428.post) = 'Deviations'
gene12428.post$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene12428.post$se = m.model$result$PosteriorSD[12428,]

p.orig = ggplot(gene12428, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + 
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) +
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('TRIM26 original log 2 fold change' )

p.post = ggplot(gene12428.post, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + 
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('TRIM26 mash')

ggarrange(p.orig, p.post, ncol=2, nrow=1, heights = c(5, 5), widths = c(5,5))
```

