---
title: "Expression 18 Mash - EZ"
author: "Yuxin Zou"
date: 2020-12-9
output: 
  html_document:
    code_folding: hide
---

```{r packages}
library(limma); library(mashr); library(plyr);
library(flashr); library(colorRamps); library(corrplot)
```

Implement `mash` on John's data at time point 18.

```{r data}
data = readRDS('../output/Expression_john.rds')
beta_class = data$beta
sebeta_class = data$sebeta
df_class = data$df
```

# Mash

Set mash data object
```{r mash.data}
mash.data = mash_set_data(Bhat = beta_class, Shat = sebeta_class, df = df_class, alpha=1)
mash.data.L = mash_update_data(mash.data, ref = 'none')
```

Top genes:
```{r top genes}
# find strong genes
m.1by1 = mash_1by1(mash.data.L, alpha=1)
strong = get_significant_results(m.1by1)

data.strong = mash_set_data(Bhat = beta_class[strong,], Shat = sebeta_class[strong,], df = df_class[strong,], alpha=1)
data.strong.L = mash_update_data(data.strong, ref = 'none')
```

## Estimate covariance structures:

Data Driven:

Flash:
```{r flash, message=FALSE}
flash.data = flash_set_data(data.strong.L$Bhat)
fmodel = flash(flash.data, greedy = TRUE, backfit = TRUE)
```

```{r flash plot}
factors = flash_get_ldf(fmodel)$f
row.names(factors) = colnames(beta_class)[2:9]
pve.order = order(fmodel$pve, decreasing = TRUE)

par(mfrow=c(1,3))
for(i in pve.order){
  barplot(factors[,i], main=paste0('Factor ',i, ' pve= ', round(fmodel$pve[i],3)), las=2, cex.names = 0.7)
}
par(mfrow=c(1,1))
```

```{r cov from flash}
Factors = fmodel$ldf$f
Flash_res = flash_get_fitted_values(fmodel)
U.Flash = c(mashr:::cov_from_factors(t(as.matrix(Factors)), "Flash"), 
            list("tFlash" = t(Flash_res) %*% Flash_res / nrow(Flash_res)))
```

PCA:
```{r pca}
U.pca = cov_pca(data.strong.L, 3)
```

Canonical
```{r Canonical}
U.c = cov_canonical(data.strong.L)
```

Extreme Deconvolution
```{r ED, eval=FALSE}
X.center = apply(data.strong.L$Bhat, 2, function(x) x - mean(x))
U.dd = c(U.pca, U.Flash, list('XX' = t(X.center) %*% X.center / nrow(X.center)))

saveRDS(cov_ed(data.strong.L, U.dd), '../output/Expression18_EZ_ED.rds')
```

## Mash model:
```{r}
U.ed = readRDS('../output/Expression18_EZ_ED.rds')
```

```{r mash, eval=FALSE}
saveRDS(mash(mash.data.L, c(U.c, U.ed)), '../output/Expression18_EZ_mash_model.rds') 
```

# Result
```{r}
mash.model = readRDS('../output/Expression18_EZ_mash_model.rds')
```

The log-likelihood of fit is
```{r} 
get_loglik(mash.model)
```
Here is a plot of weights learned:
```{r}
options(repr.plot.width=12, repr.plot.height=4)
barplot(get_estimated_pi(mash.model), las = 2, cex.names = 0.7)
```

Check tPCA covariance matrix

```{r tPCA}
x           <- mash.model$fitted_g$Ulist[["ED_tPCA"]]
colnames(x) <- colnames(beta_class)[2:9]
rownames(x) <- colnames(x)
corrplot(x, method='color', cl.lim=c(0,1), type='upper', addCoef.col = "black", tl.col="black", tl.srt=45)
```

```{r tPCA pca}
layout(matrix(c(1,2,3,4), 2, 2, byrow=TRUE))
svd.out = svd(mash.model$fitted_g$Ulist[["ED_tPCA"]])
v = svd.out$v
colnames(v) = colnames(beta_class)[-1]
rownames(v) = colnames(v)
options(repr.plot.width=10, repr.plot.height=5)
for (j in 1:4)
  barplot(v[,j]/v[,j][which.max(abs(v[,j]))], cex.names = 0.7,
          las = 2, main = paste0("EigenVector ", j, " for ED_tPCA"))
```

There are `r length(get_significant_results(mash.model))` diferentially expressed genes.

Check pairwise sharing by magnitude and sign:
```{r sharingMagSign}
x = get_pairwise_sharing(mash.model)
x[x > 1]    <- 1
x[x < -1]   <- -1
colnames(x) <- colnames(beta_class)[-1]
rownames(x) <- colnames(x)
corrplot.mixed(x, tl.pos="d",upper='color', cl.lim=c(0,1), upper.col=colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF",
                               "#E0F3F8","#91BFDB","#4575B4")))(40),
               tl.cex=1.2)
```
Check pairwise sharing by magnitude:
```{r shareMag}
x = get_pairwise_sharing(mash.model, FUN = abs)
x[x > 1]    <- 1
x[x < -1]   <- -1
colnames(x) <- colnames(beta_class)[-1]
rownames(x) <- colnames(x)
corrplot.mixed(x, tl.pos="d",upper='color', cl.lim=c(0,1), upper.col=colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF",
                               "#E0F3F8","#91BFDB","#4575B4")))(40),
               tl.cex=1.2)
```

Check pairwise sharing by sign:
```{r shareSign}
x = get_pairwise_sharing(mash.model, factor=0)
x[x > 1]    <- 1
x[x < -1]   <- -1
colnames(x) <- colnames(beta_class)[-1]
rownames(x) <- colnames(x)
corrplot.mixed(x, tl.pos="d",upper='color', cl.lim=c(0,1), upper.col=colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF",
                               "#E0F3F8","#91BFDB","#4575B4")))(40),
               tl.cex=1.2)
```
# Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
