---
title: "Expression 18 Mash - reprocess18 - EE - joint mean"
author: "Yuxin Zou"
date: 2021-2-25
output: 
  html_document:
    code_folding: hide
---

```{r packages}
library(limma); library(mashr); library(plyr);
library(flashr); library(colorRamps); library(corrplot)
library(ggplot2); library(ggpubr)
```

Implement `mash` on John's data at time point 18.

```{r data}
data = readRDS('../output/Expression_reprocesstime18_joint.rds')
beta_class = data$beta
sebeta_class = data$sebeta

mash.data = mash_set_data(beta_class, sebeta_class)
mash.data.L = mash_update_data(mash.data, ref='none')
```

Top genes:
```{r top genes}
# find strong genes
m.1by1 = mash_1by1(mash.data.L, alpha=0)
strong = get_significant_results(m.1by1)

data.strong = mash_set_data(Bhat = beta_class[strong,], Shat = sebeta_class[strong,])
data.strong.L = mash_update_data(data.strong, ref = 'none')
```
There are `r length(strong)` strong genes for at least one bacteria using `ash` condition-by-condition. We use DE genes identified in John's paper as strong set:
```{r}
john = readRDS('../data/JohnPaper18.rds')
topgenes = names(john$motif)[john$motif!='non-DE']
strong = which(rownames(beta_class) %in% topgenes)
data.strong = mash_set_data(Bhat = beta_class[strong,], Shat = sebeta_class[strong,])
data.strong.L = mash_update_data(data.strong, ref = 'none')
```

## Estimate covariance structures

Data Driven:

Flash:
```{r flash, message=FALSE}
flash.data = flash_set_data(data.strong.L$Bhat)
fmodel = flash(flash.data, greedy = TRUE, backfit = TRUE)
```

```{r flash plot}
factors = flash_get_ldf(fmodel)$f
row.names(factors) = colnames(beta_class)[2:9]
pve.order = order(fmodel$pve, decreasing = TRUE)

par(mfrow=c(2,3))
for(i in pve.order){
  barplot(factors[,i], main=paste0('Factor ',i, ' pve= ', round(fmodel$pve[i],3)), las=2, cex.names = 0.7)
}
par(mfrow=c(1,1))
```

```{r cov from flash}
Factors = fmodel$ldf$f
Flash_res = flash_get_fitted_values(fmodel)
U.Flash = c(mashr:::cov_from_factors(t(as.matrix(Factors)), "Flash"), 
            list("tFlash" = t(Flash_res) %*% Flash_res / nrow(Flash_res)))
```

PCA:
```{r pca}
U.pca = cov_pca(data.strong.L, 3)
```

Canonical
```{r Canonical}
U.c = cov_canonical(data.strong.L)
```

Extreme Deconvolution
```{r ED, eval=FALSE}
X.center = apply(data.strong.L$Bhat, 2, function(x) x - mean(x))
U.dd = c(U.pca, U.Flash, list('XX' = t(X.center) %*% X.center / nrow(X.center)))

saveRDS(cov_ed(data.strong.L, U.dd), '../output/Expression18_reprocesstime18_EE_jointmean_ED.rds')
```

## Mash model

```{r}
U.ed = readRDS('../output/Expression18_reprocesstime18_EE_jointmean_ED.rds')
```

```{r mash, eval=FALSE}
U.dd.diag = c(U.ed[c('ED_tPCA', 'ED_tFlash', 'ED_XX')], lapply(U.ed[c('ED_PCA_1','ED_PCA_2','ED_PCA_3', paste0('ED_Flash_',1:6))], function(U) U + 0.01*diag(8)))
mash.model = mash(mash.data.L, c(U.c, U.dd.diag))
saveRDS(mash.model, '../output/Expression18_reprocesstime18_EE_jointmean_mash_model.rds') 
```

## Result

```{r}
mash.model = readRDS('../output/Expression18_reprocesstime18_EE_jointmean_mash_model.rds')
```

The log-likelihood of fit is
```{r} 
get_loglik(mash.model)
```
Here is a plot of weights learned:
```{r}
options(repr.plot.width=12, repr.plot.height=4)
barplot(get_estimated_pi(mash.model), las = 2, cex.names = 0.7)
```

There are `r length(get_significant_results(mash.model))` differentially expressed genes.

Check tPCA covariance matrix

```{r tPCA}
x           <- mash.model$fitted_g$Ulist[["ED_tPCA"]]
colnames(x) <- colnames(beta_class)[2:9]
rownames(x) <- colnames(x)
corrplot(x, method='color', cl.lim=c(0,1), type='upper', addCoef.col = "black", tl.col="black", tl.srt=45, tl.cex = 0.7)
```

```{r tPCA pca}
layout(matrix(c(1,2,3,4), 2, 2, byrow=TRUE))
svd.out = svd(mash.model$fitted_g$Ulist[["ED_tPCA"]])
v = svd.out$v
colnames(v) = colnames(beta_class)[-1]
rownames(v) = colnames(v)
options(repr.plot.width=10, repr.plot.height=5)
for (j in 1:4)
  barplot(v[,j]/v[,j][which.max(abs(v[,j]))], cex.names = 0.7,
          las = 2, main = paste0("EigenVector ", j, " for ED_tPCA"))
```

Check pairwise sharing by magnitude and sign:
```{r sharingMagSign}
x = get_pairwise_sharing(mash.model)
x[x > 1]    <- 1
x[x < -1]   <- -1
colnames(x) <- colnames(beta_class)[-1]
rownames(x) <- colnames(x)
corrplot.mixed(x, tl.pos="d",upper='color', cl.lim=c(0,1), upper.col=colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF",
                               "#E0F3F8","#91BFDB","#4575B4")))(40),
               tl.cex=1.2)
```

Check pairwise sharing by magnitude:
```{r shareMag}
x = get_pairwise_sharing(mash.model, FUN = abs)
x[x > 1]    <- 1
x[x < -1]   <- -1
colnames(x) <- colnames(beta_class)[-1]
rownames(x) <- colnames(x)
corrplot.mixed(x, tl.pos="d",upper='color', cl.lim=c(0,1), upper.col=colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF",
                               "#E0F3F8","#91BFDB","#4575B4")))(40),
               tl.cex=1.2)
```

Check pairwise sharing by sign:
```{r shareSign}
x = get_pairwise_sharing(mash.model, factor=0)
x[x > 1]    <- 1
x[x < -1]   <- -1
colnames(x) <- colnames(beta_class)[-1]
rownames(x) <- colnames(x)
corrplot.mixed(x, tl.pos="d",upper='color', cl.lim=c(0,1), upper.col=colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF",
                               "#E0F3F8","#91BFDB","#4575B4")))(40),
               tl.cex=1.2)
```

## Genes

Yers-Salm cluster in Cormotif, not significant in mash.
```{r YSgene1, fig.height=5, fig.width=10}
# before
gene1177 = data.frame(mash.data.L$Bhat[1177,])
colnames(gene1177) = 'Deviations'
gene1177$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene1177$se = mash.data.L$Shat[1177,]
gene1177$Deviations[gene1177$se == 0] = NA
# after
gene1177.post = data.frame(mash.model$result$PosteriorMean[1177,])
colnames(gene1177.post) = 'Deviations'
gene1177.post$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene1177.post$se = mash.model$result$PosteriorSD[1177,]

p.orig = ggplot(gene1177, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + ylim(-3,4) +
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('ENSG00000079335 original' )

p.post = ggplot(gene1177.post, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + ylim(-3,4) +
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('ENSG00000079335 mash')

ggarrange(p.orig, p.post, ncol=2, nrow=1, heights = c(5, 5), widths = c(5,5))
```

Yers-Salm cluster in Cormotif, not significant in mash.
```{r YSgene2, fig.height=5, fig.width=10}
# before
gene7637 = data.frame(mash.data.L$Bhat[7637,])
colnames(gene7637) = 'Deviations'
gene7637$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene7637$se = mash.data.L$Shat[7637,]
gene7637$Deviations[gene7637$se == 0] = NA
# after
gene7637.post = data.frame(mash.model$result$PosteriorMean[7637,])
colnames(gene7637.post) = 'Deviations'
gene7637.post$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene7637.post$se = mash.model$result$PosteriorSD[7637,]

p.orig = ggplot(gene7637, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + ylim(-10,8) +
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('ENSG00000158488 original' )

p.post = ggplot(gene7637.post, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + ylim(-10, 8) +
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('ENSG00000158488 mash')

ggarrange(p.orig, p.post, ncol=2, nrow=1, heights = c(5, 5), widths = c(5,5))
```

Yers-Salm cluster in Cormotif, not significant in mash.
```{r YSgene3, fig.height=5, fig.width=10}
# before
gene1631 = data.frame(mash.data.L$Bhat[1631,])
colnames(gene1631) = 'Deviations'
gene1631$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene1631$se = mash.data.L$Shat[1631,]
gene1631$Deviations[gene1631$se == 0] = NA
# after
gene1631.post = data.frame(mash.model$result$PosteriorMean[1631,])
colnames(gene1631.post) = 'Deviations'
gene1631.post$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene1631.post$se = mash.model$result$PosteriorSD[1631,]

p.orig = ggplot(gene1631, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + ylim(-5,5) +
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('ENSG00000095380 original' )

p.post = ggplot(gene1631.post, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + ylim(-5, 5) +
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('ENSG00000095380 mash')

ggarrange(p.orig, p.post, ncol=2, nrow=1, heights = c(5, 5), widths = c(5,5))
```

All in Cormofit, not significant in mash.
```{r allgene1, fig.height=5, fig.width=10}
# before
gene93 = data.frame(mash.data.L$Bhat[93,])
colnames(gene93) = 'Deviations'
gene93$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene93$se = mash.data.L$Shat[93,]
gene93$Deviations[gene93$se == 0] = NA
# after
gene93.post = data.frame(mash.model$result$PosteriorMean[93,])
colnames(gene93.post) = 'Deviations'
gene93.post$Group = factor(c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'), levels = c('Rv', 'Rvplus', 'GC', 'BCG', 'Smeg', 'Yers', 'Salm', 'Staph'))
gene93.post$se = mash.model$result$PosteriorSD[93,]

p.orig = ggplot(gene93, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + ylim(-6,4) +
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) +
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('ENSG00000006015 original' )

p.post = ggplot(gene93.post, aes(y = Deviations, x = Group, color=Group)) + 
  geom_point() + ylim(-6, 4) +
  geom_errorbar(aes(ymin=Deviations-1.96*se, ymax=Deviations+1.96*se), width=0.4) + 
  geom_hline(yintercept = 0, linetype="dotted", color = "grey", size=1) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),axis.line = element_line(colour = "black"), legend.position = "none", plot.title = element_text(size = 10), axis.text=element_text(size=10)) + coord_flip() + ggtitle('ENSG00000006015 mash')

ggarrange(p.orig, p.post, ncol=2, nrow=1, heights = c(5, 5), widths = c(5,5))
```

## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
