---
title: "Count - Expression 18"
author: "Yuxin Zou"
date: 2021-2-25
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

```{r, echo=FALSE}
# TEMPORARY.
knitr::opts_chunk$set(eval = TRUE)
```

The type of bacteria (bact), and the time post-infection (time) are treated as fixed effects. The individual (ind) is treated as fixed effect.

The covariates RIN (rin) and extraction batch (extr) are included in the model to correct for these confounding effects.

```{r packages}
library("limma")
library("edgeR")
library('lme4')
library('car')
library(dplyr)
```

Read in the data of counts (row = sample, column = feature).

```{r data}
full <- read.table("../data/counts_per_sample.txt", header = TRUE,
                   stringsAsFactors = FALSE)
full <- full[order(full$dir), ]
rownames(full) <- paste(full$dir, full$ind, full$bact, full$time, sep = ".")
counts <- t(full[, grep("ENSG", colnames(full))])
# Filter lowly expressed genes
counts <- counts[rowSums(cpm(counts) > 1) > 6, ]
```

The first `r sum(!grepl("ENSG", colnames(full)))` columns are meta-data about each sample, and the remaining are the gene counts.

Only the `r nrow(counts)` genes which have at least one read per million (pre-normalization with TMM) in at least six of the `r nrow(full)` samples are kept.

Prepare the factors for use in the linear model.

```{r annotation}
groups <- full[, c("ind", "bact", "time", "extr", "rin")]
groups$bact <- gsub("\\+", "plus", groups$bact)
groups$ind <- factor(groups$ind)
groups$time <- factor(groups$time, levels = c(4, 18, 48))
groups$bact <- factor(groups$bact, levels = c("none", "Rv", "Rvplus", "GC",
                                              "BCG", "Smeg", "Yers", "Salm",
                                              "Staph"))
groups$extr <- factor(groups$extr)
head(groups)
```

* Normalize with TMM (`calcNormFactors`).

```{r TMM}
# time 18
idx = which(groups$time == 18)
groups = groups[idx,]
counts = counts[, idx]
y <- DGEList(counts)
y <- calcNormFactors(y)
```

* Create the model using a group means parametrization
```{r model}
contrasts(groups$extr) = contr.Sum(levels(factor(groups$extr)))
contrasts(groups$ind) = contr.Sum(levels(groups$ind))
design <- model.matrix(~ 0 + bact + ind + extr + rin, data = groups)
colnames(design) <- gsub("bact", "", colnames(design))
colnames(design) <- gsub(":", ".", colnames(design))
colnames(design) <- gsub("ind", "", colnames(design))
print(colnames(design))
```

* Transfer with `voom`
```{r voom}
v = voom(y, design)
```

* limma
```{r limma}
fit = lmFit(v, design)
fitB = eBayes(fit)
```

```{r assign}
beta_class = fitB$coefficients[,1:9]
sebeta_class = fitB$stdev.unscaled[, 1:9] * sqrt(fitB$s2.post)
t_class = fitB$t[, 1:9]
df_class = fitB$df.total
saveRDS(list(beta = beta_class, sebeta = sebeta_class, t = t_class, df = df_class, groups = groups),
        '../output/Expression_reprocess18_joint.rds')
```


# Session information

<!-- Insert the session information into the document -->
```{r session-info}
```

