---
title: "Expression 18 Mash - EE - joint difference"
author: "Yuxin Zou"
date: 2020-12-9
output: 
  html_document:
    code_folding: hide
---

```{r packages}
library(limma); library(mashr); library(plyr);
library(flashr); library(colorRamps); library(corrplot)
library(ggplot2); library(ggpubr)
```

Implement `mash` on John's data at time point 18.

```{r data}
data = readRDS('../output/Expression_john_diff_joint.rds')
beta_class = data$beta
sebeta_class = data$sebeta
df_class = data$df
```

Set mash data object
```{r mash.data}
mash.data = mash_set_data(Bhat = beta_class, Shat = sebeta_class, df=df_class)
```

Top genes:
```{r top genes}
# find strong genes
m.1by1 = mash_1by1(mash.data, alpha=0)
strong = get_significant_results(m.1by1)
```
There are `r length(strong)` strong genes for at least one bacteria using `ash` condition-by-condition.

## Use all strong genes

```{r}
data.strong = mash_set_data(Bhat = beta_class[strong,], Shat = sebeta_class[strong,], df=df_class[strong,])
```

Flash:
```{r, eval=FALSE}
flash.data = flash_set_data(data.strong$Bhat)
fmodel = flash(flash.data, greedy = TRUE, backfit = TRUE)
saveRDS(fmodel, '../output/Expression18_EE_jointdiff_flash.rds')
```

```{r cov from flash}
fmodel = readRDS('../output/Expression18_EE_jointdiff_flash.rds')
Factors = fmodel$ldf$f
Flash_res = flash_get_fitted_values(fmodel)
U.Flash = c(mashr:::cov_from_factors(t(as.matrix(Factors)), "Flash"), 
            list("tFlash" = t(Flash_res) %*% Flash_res / nrow(Flash_res)))
```

```{r}
factors = flash_get_ldf(fmodel)$f
row.names(factors) = colnames(beta_class)
pve.order = order(fmodel$pve, decreasing = TRUE)

par(mfrow=c(2,3))
for(i in pve.order){
  barplot(factors[,i], main=paste0('Factor ',i, ' pve= ', round(fmodel$pve[i],3)), las=2, cex.names = 0.7)
}
par(mfrow=c(1,1))
```

PCA:
```{r pca}
U.pca = cov_pca(data.strong, 3)
```

Canonical:
```{r Canonical}
U.c = cov_canonical(data.strong)
```

ED:
```{r ED, eval=FALSE}
X.center = apply(data.strong$Bhat, 2, function(x) x - mean(x))
U.dd = c(U.pca, U.Flash, list('XX' = t(X.center) %*% X.center / nrow(X.center)))
saveRDS(cov_ed(data.strong, U.dd), '../output/Expression18_EE_jointdiff_ED.rds')
```

```{r mash, eval=FALSE}
U.ed = readRDS('../output/Expression18_EE_jointdiff_ED.rds')
saveRDS(mash(mash.data, c(U.c, U.ed)), '../output/Expression18_jointdiff_mash_model.rds') 
```

```{r}
mash.model = readRDS('../output/Expression18_jointdiff_mash_model.rds') 
```
There are `r length(get_significant_results(mash.model))` DE genes.


## Use random subset of strong genes

We randomly sample 5000 strong genes to learn data driven patterns.

```{r}
set.seed(1234)
idx = sample(strong, 5000, replace = FALSE)
data.strong = mash_set_data(Bhat = beta_class[idx,], Shat = sebeta_class[idx,], df=df_class[idx,])
```

We randomly sample 7000 genes to fit mash model.

```{r}
set.seed(3456)
idx = sample(1:nrow(beta_class), 7000, replace = FALSE)
data.random = mash_set_data(Bhat = beta_class[idx,], Shat = sebeta_class[idx,], df=df_class[idx,])
```

Flash:
```{r, eval=FALSE}
flash.data = flash_set_data(data.strong$Bhat)
fmodel = flash(flash.data, greedy = TRUE, backfit = TRUE)
saveRDS(fmodel, '../output/Expression18_EE_jointdiff_subset_flash.rds')
```

```{r subset pca}
U.pca = cov_pca(data.strong, 3)
```

Canonical:
```{r subset Canonical}
U.c = cov_canonical(data.strong)
```

ED:
```{r subset ED, eval=FALSE}
X.center = apply(data.strong$Bhat, 2, function(x) x - mean(x))
U.dd = c(U.pca, U.Flash, list('XX' = t(X.center) %*% X.center / nrow(X.center)))
saveRDS(cov_ed(data.strong, U.dd), '../output/Expression18_EE_jointdiff_subset_ED.rds')
```

```{r subset mash, eval=FALSE}
U.ed = readRDS('../output/Expression18_EE_jointdiff_subset_ED.rds')
m.model = mash(data.random, c(U.c, U.ed), outputlevel = 1)
m.model$result = mash_compute_posterior_matrices(m.model, mash.data)
saveRDS(m.model, '../output/Expression18_jointdiff_subset_mash_model.rds') 
```

```{r}
mash.model = readRDS('../output/Expression18_jointdiff_subset_mash_model.rds') 
```
There are `r length(get_significant_results(mash.model))` DE genes.


