---
title: "Expression simulation under null"
author: "Yuxin Zou"
date: 2021-2-19
output: 
  html_document:
    code_folding: hide
---

```{r}
library(limma)
library(ashr)
library(mashr)
```

The data includes the gene expression data (table_s1) of 12,728 genes from 54 samples infected by 8 different types of bacteria and one non-infected control, measured at 18 hours post-infection. 

## Same permutation for all gene

We randomly permute the samples with respect to the condition labels so that there no longer exists association between gene expression levels and conditions.

```{r}
dat = readRDS('../output/Expression_data.rds')
set.seed(123)

perm = sample(1:54, 54)
d1 = dat$data[,perm]
colnames(d1) = colnames(dat$data)
  
fit1=lmFit(object = d1, dat$design_diff)
fit1<-eBayes(fit1)
b1 = fit1$coefficients[,-1]
s1 <- fit1$stdev.unscaled[,-1] * sqrt(fit1$s2.post)
t1 <- fit1$t[,-1]
df1 <- fit1$df.total

fit2=lmFit(object = d1, dat$design_mean)
fit2<-eBayes(fit2)
b2 = fit2$coefficients
s2 <- fit2$stdev.unscaled * sqrt(fit2$s2.post)
t2 <- fit2$t
df2 <- fit2$df.total
```

Fit ash for each condition:
```{r}
m1.data = mash_set_data(b1, s1)
m1.1by1 = mash_1by1(m1.data)
fpash = colSums(m1.1by1$result$lfsr < 0.05)
```

The false positive rates using ash:
```{r}
fpash/nrow(d1)
```

Fit mash on DE data:
```{r}
strong1 = get_significant_results(m1.1by1, 0.05)
U1.pca = cov_pca(m1.data, 5, subset=strong1)
U1.ed = cov_ed(m1.data, U1.pca, subset=strong1)
U1.c = cov_canonical(m1.data)
m1.fit.diff = mash(m1.data, c(U1.c,U1.ed))
fpmash.diff = length(get_significant_results(m1.fit.diff))
```

The FPR is `r fpmash.diff/nrow(b1)`. We check estimated weights:
```{r}
barplot(get_estimated_pi(m1.fit.diff), las=2, cex.names = 0.7)
```

The ED priors already adjusted with a small diagonal elements. We add a small number on the diagonal of cannonical priors:
```{r}
U1.c.diag = lapply(U1.c, function(U) U + 0.01 * diag(8))
m1.fit.diff.diag = mash(m1.data, c(U1.c.diag, U1.ed))
fpmash.diff.diag = length(get_significant_results(m1.fit.diff.diag))
```

The FPR is `r fpmash.diff.diag/nrow(b1)`. We check estimated weights:
```{r}
barplot(get_estimated_pi(m1.fit.diff.diag), las=2, cex.names = 0.7)
```

Fit mash common baseline:
```{r}
m2.data = mash_set_data(b2, s2)
m2.data.L = mash_update_data(m2.data, ref='none')
m2.1by1 = mash_1by1(m2.data.L)
strong2 = get_significant_results(m2.1by1, 0.05)
U2.pca = cov_pca(m2.data.L, 5, subset=strong2)
U2.ed = cov_ed(m2.data.L, U2.pca, subset=strong2)
U2.c = cov_canonical(m2.data.L)

m2.fit.mean = mash(m2.data.L, c(U2.c,U2.ed))
fpmash.mean = length(get_significant_results(m2.fit.mean))
```

The FPR is `r fpmash.mean/nrow(b2)`. We check estimated weights:
```{r}
barplot(get_estimated_pi(m2.fit.mean), las=2, cex.names = 0.7)
```

Add a small number on the diagonal of priors:
```{r}
U2.c.diag = lapply(U2.c, function(U) U + 0.01 * diag(8))
m2.fit.mean.diag = mash(m2.data.L, c(U2.c.diag, U2.ed))
fpmash.mean.diag = length(get_significant_results(m2.fit.mean.diag))
```

The FPR is `r fpmash.mean.diag/nrow(b2)`. We check estimated weights:
```{r}
barplot(get_estimated_pi(m2.fit.mean.diag), las=2, cex.names = 0.7)
```

If we exclude data-driven priors,
```{r}
m2.fit.mean.c = mash(m2.data.L, U2.c)
fpmash.mean.c = length(get_significant_results(m2.fit.mean.c))
```

The FPR is `r fpmash.mean.c/nrow(b2)`. We check estimated weights:
```{r}
barplot(get_estimated_pi(m2.fit.mean.c), las=2, cex.names = 0.7)
```

## Different permutation for each gene

We use independent and different permutation to each gene.

```{r}
dat = readRDS('../output/Expression_data.rds')
set.seed(123)
d2 = dat$data
for(i in 1:nrow(dat$data)){
  perm = sample(1:54, 54)
  d2[i,] = d2[i,perm]
}

fit3=lmFit(object = d2, dat$design_diff)
fit3<-eBayes(fit3)
b3 = fit3$coefficients[,-1]
s3 <- fit3$stdev.unscaled[,-1] * sqrt(fit3$s2.post)
t3 <- fit3$t[,-1]
df3 <- fit3$df.total

fit4=lmFit(object = d2, dat$design_mean)
fit4<-eBayes(fit4)
b4 = fit4$coefficients
s4 <- fit4$stdev.unscaled * sqrt(fit4$s2.post)
t4 <- fit4$t
df4 <- fit4$df.total
```

Fit ash for each condition:
```{r}
m3.data = mash_set_data(b3, s3)
m3.1by1 = mash_1by1(m3.data)
fpash.g = colSums(m3.1by1$result$lfsr < 0.05)
```

The false positive rates using ash:
```{r}
fpash.g/nrow(d2)
```

Fit mash on DE data:
```{r}
U3.c = cov_canonical(m3.data)
m3.fit.diff = mash(m3.data, U3.c)
fpmash.diff.g = length(get_significant_results(m3.fit.diff))
```

The FPR is `r fpmash.diff.g/nrow(d2)`. We check estimated weights:
```{r}
barplot(get_estimated_pi(m3.fit.diff), las=2, cex.names = 0.7)
```

Fit mash common baseline:
```{r}
m4.data = mash_set_data(b4, s4)
m4.data.L = mash_update_data(m4.data, ref='none')
m4.1by1 = mash_1by1(m4.data.L)
U4.c = cov_canonical(m4.data.L)
m4.fit.mean = mash(m4.data.L, U4.c)
fpmash.mean.g = length(get_significant_results(m4.fit.mean))
```

The FPR is `r fpmash.mean.g/nrow(b2)`. We check estimated weights:
```{r}
barplot(get_estimated_pi(m4.fit.mean), las=2, cex.names = 0.7)
```

